Classification structurée pour l'apprentissage par renforcement

### 2012-09-14 klein2012structured_RIA/scirl_ria.tex
Section mise en pratique traduite

\section{A Practical Approach}
\label{sec:practicalApproach}

\subsection{Estimating the Expert Feature Expectation}
\label{subsec:practicalApproach:muE}

SCIRL relies on an estimate $\hat{\mu}^{\pi_E}$ of the expert
feature expectation. Basically, this is a policy evaluation problem.
An already made key observation is that each component of
$\mu^{\pi_E}$ is the action-value function of $\pi_E$ for a reward
function $\phi_i$: $\mu_i^{\pi_E}(s,a) = Q^{\pi_E}_{\phi_i}(s,a) =
[T^a_{\phi_i} v^{\pi_E}_{\phi_i}](s)$. We briefly review its exact
computation et possible estimation approaches, et consider
possible heuristics.
%possibly considering some heuristics.

If the model is known, the feature expectation can be computed
explicitly. Let $\Phi\in\mathbb{R}^{|\s|\times p}$ be the feature
matrix whose rows contain the feature vectors $\phi(s)^\top$ for all
$s\in\s$.
%
%(equivalently, whose columns contain the basis functions $\phi_i$
%for $1\leq i\leq p$).
%
For a fixed $a\in A$, let $\Mu^{\pi_E}_a \in\mathbb{R}^{|\s|\times
p}$ be the feature expectation matrix whose rows are the expert
feature vectors, that is $(\mu^{\pi_E}(s,a))^\top$ for any $s\in\s$.
With these notations, we have $\Mu_a^{\pi_E} = \Phi + \gamma
P_a(I-\gamma P_{\pi_E})^{-1} \Phi$.
%\begin{equation}
%  \forall a\in\A,\; \Mu_a^{\pi_E} = \Phi + \gamma P_a(I-\gamma
%  P_{\pi_E})^{-1} \Phi.
%\end{equation}
%Therefore, knowing the model (the dynamic), the expert feature
%expectation can be easily computed.
Moreover, the related computational cost is the same order of
magnitude as evaluating a single policy (as the costly part,
computing $(I-\gamma P_{\pi_E})^{-1}$, is shared by all components).

If the model is unknown, any temporal difference learning algorithm
can be used to estimate the expert feature
expectation~\cite{Klein:2011}, as LSTD (Least-Squares Temporal
Differences)~\cite{Bradtke:1996}. Let $\psi:\s\times \A \rightarrow
\mathbb{R}^d$ be a feature vector composed of $d$ basis functions
$\psi_i \in\mathbb{R}^{\s\times \A}$. Each component $\mu_i^{\pi_E}$
of the expert feature expectation is parameterized by a vector
$\xi_i\in\mathbb{R}^d$: $\mu_i^{\pi_E}(s,a)\approx \xi_i^\top
\psi(s,a)$. Assume that we have a training set
$\{(s_i,a_i,s'_i,a'_i=\pi_E(s'_i))_{1\leq i \leq M}\}$ with actions
$a_i$ not necessarily sampled according to policy $\pi_E$
(\textit{e.g.}, this may be obtained by sampling trajectories
according to an expert-based $\epsilon$-greedy policy), the aim
being to have a better variability of tuples (non-expert actions
should be tried). Let $\tilde{\Psi}\in \mathbb{R}^{M\times d}$
(resp. $\tilde{\Psi}'$) be the feature matrix whose rows are the
feature vectors $\psi(s_i,a_i)^\top$ (resp. $\psi(s'_i,a'_i)^\top$).
Let also $\tilde{\Phi}\in \mathbb{R}^{M\times p}$ be the feature
matrix whose rows are the reward's feature vectors $\phi(s_i)^\top$.
Finally, let $\Xi =
\begin{bmatrix}   \xi_1 & \dots & \xi_p
\end{bmatrix}\in\mathbb{R}^{d\times p}$ be the matrix of all
parameter vectors. Applying LSTD to each component of the feature
expectation gives the LSTD-$\mu$ algorithm~\cite{Klein:2011}: $\Xi =
(\tilde{\Psi}^\top(\tilde{\Psi} - \gamma
\tilde{\Psi}'))^{-1}\tilde{\Psi}^\top \tilde{\Phi}$ and
$\hat{\mu}^{\pi_E}(s,a) = \Xi^\top \psi(s,a)$.
%\begin{equation}
%  \Xi = \left(\tilde{\Psi}^\top(\tilde{\Psi} - \gamma
%  \tilde{\Psi}')\right)^{-1}\tilde{\Psi}^\top \tilde{\Phi}
%  \text{ et } \hat{\mu}^{\pi_E}(s,a) = \Xi^\top \psi(s,a).
%\end{equation}
As for the exact case, the costly part (computing the inverse
matrix) is shared by all feature expectation components, the
computational cost is reasonable (same order as LSTD).

%
Provided a simulator et the ability to sample according to the
expert policy, the expert feature expectation may also be estimated
using Monte-Carlo rollouts for a given state-action pair (as noted
in Sec.~\ref{subsec:scirl:analysis}, $\hat{\mu}^{\pi_E}$ need only
be known on $(s_i,a)_{1\leq i\leq N,a\in\A}$). Assuming that $K$
trajectories are sampled for each required state-action pair, this
method would require $KN|\A|$ rollouts.

In order to have a small error $\bar{\epsilon}_Q$, one may learn
using transitions whose starting state is sampled according to
$\rho_E$ et whose actions are uniformly distributed. However, it
may happen that only transitions of the expert are available: $\T =
\{(s_i,a_i=\pi_E(s_i),s'_i)_{1\leq i \leq N}\}$. If the state-action
couples $(s_i,a_i)$ may be used to feed the classifier, the
transitions $(s_i,a_i,s'_i)$ are not enough to provide an accurate
estimate of the feature expectation. In this case, we can still
expect an accurate estimate of $\mu^{\pi_E}(s,\pi_E(s))$, but there
is little hope for $\mu^{\pi_E}(s,a\neq\pi_E(s))$. However, one can
still rely on some heuristic; this does not fit the analysis of
Sec.~\ref{subsec:scirl:analysis}, but it can still provide good
experimental results, as illustrated in Sec.~\ref{sec:experiments}.

We propose such a heuristic. Assume that only data $\T$ is available
and that we use it to provide an (accurate) estimate
$\hat{\mu}^{\pi_E}(s,\pi_E(s))$ (this basically means estimating a
value function instead of an action-value function as described
above). We may adopt an optimistic point of view by assuming that
applying a non-expert action just delays the effect of the expert
action. More formally, we associate to each state $s$ a virtual
state $s_\text{v}$ for which $p(.|s_\text{v},a)=p(.|s,\pi_E(s))$ for
any action $a$ et for which the reward feature expectation is the
null vector, $\phi(s_v) = 0$. In this case, we have
$\mu^{\pi_E}(s,a\neq\pi_E(s)) = \gamma \mu^{\pi_E}(s,\pi_E(s))$.
Applying this idea to the available estimate (recalling that the
classifiers only requires evaluating $\hat{\mu}^{\pi_E}$ on
$(s_i,a)_{1\leq i\leq N,a\in \A}$) provides the proposed heuristic:
for $1\leq i\leq N$, $\hat{\mu}^{\pi_E}(s_i,a\neq a_i) = \gamma
\hat{\mu}^{\pi_E}(s_i,a_i)$.
%\begin{equation}
%  \forall 1\leq i \leq N,\;
%  \hat{\mu}^{\pi_E}(s_i,a\neq a_i) = \gamma
%  \hat{\mu}^{\pi_E}(s_i,a_i).
%  \label{eq:heuristic}
%\end{equation}

We may even push this idea further, to get the simpler estimate of
the expert feature expectation (but with the weakest guarantees).
Assume that the set $\T$ consists of one long trajectory, that is
$s'_i = s_{i+1}$ (thus $\T =
\{s_1,a_1,s_2,\dots,s_{N-1},a_{N-1},s_N,a_N\}$). We may estimate
$\mu^{\pi_E}(s_i,a_i)$ using the single rollout available in the
training set et use the proposed heuristic for other actions:
\begin{equation}
  \forall 1\leq i \leq N,\; \hat{\mu}^{\pi_E}(s_i,a_i) =
  \sum_{j=i}^N \gamma^{j-i}\phi(s_j) \text{ et }
  \hat{\mu}^{\pi_E}(s_i,a\neq a_i) = \gamma
  \hat{\mu}^{\pi_E}(s_i,a_i).
  \label{eq:mc_plus_heuristic}
\end{equation}

To sum up, the expert feature expectation may be seen as a vector of
action-value functions (for the same policy $\pi_E$ et different
reward functions $\phi_i$). Consequently, any action-value function
evaluation algorithm may be used to estimate $\mu^\pi(s,a)$.
Depending on the available data, one may have to rely on some
heuristic to assess the feature expectation for a unexperienced
(non-expert) action. Also, this expert feature expectation estimate
is only required for training the classifier, so it is sufficient to
estimate on state-action couples $(s_i,a)_{1\leq i \leq N,a\in \A}$.
In any case, estimating $\mu^{\pi_E}$ is not harder than estimating
the action-value function of a given policy in the on-policy case,
which is much easier than computing an optimal policy for an
arbitrary reward function (as required by most of existing IRL
algorithms, see Sec.~\ref{sec:relatedWorks}).

\subsection{An Instantiation}
\label{subsec:practicalApproach:instantiation}

As stated before, any MC$^2$ algorithm may be used. Here, we choose
the structured large margin approach~\cite{Taskar:2005}. Let
$\lc:\s\times\A\rightarrow\mathbb{R}_+$ be a user-defined margin
function satisfying $\lc(s,\pi_E(s))\leq \lc(s,a)$ (here,
$\lc(s_i,a_i)=0$ et $\lc(s_i,a\neq a_i)=1$). The MC$^2$ algorithm
solves:% the following quadratic problem:
\begin{equation}
  \min_{\theta,\zeta}\frac{1}{2}\|\theta\|^2 +
  \frac{\eta}{N}\sum_{i=1}^N \zeta_i \text{~~~~s.t.~~~~} \forall i,
  \theta^\top\hat{\mu}^{\pi_E}(s_i,a_i)+\zeta_i \geq \max_a \theta^\top
  \hat{\mu}^{\pi_E}(s_i,a) + \lc(s_i,a). \label{eq:qp_taskar}
\end{equation}
Following~\cite{Ratliff:2006}, we express the equivalent hinge-loss
form (noting that the slack variables $\zeta_i$ are tight, which
allows moving the constraints in the objective function):
\begin{equation}
  J(\theta) = \frac{1}{N}\sum_{i=1}^N \max_a \theta^\top
  \hat{\mu}^{\pi_E}(s_i,a) + \lc(s_i,a) -
  \theta^\top\hat{\mu}^{\pi_E}(s_i,a_i) +
  \frac{\lambda}{2}\|\theta\|^2.
\end{equation}
This objective function is minimized using a subgradient descent.
The expert feature expectation is estimated using the scheme
described in Eq.~\eqref{eq:mc_plus_heuristic}.

>>>>

\section{Mise en pratique}
\label{sec:practicalApproach}

\subsection{Estimation de l'attribut moyen de l'expert}
\label{subsec:practicalApproach:muE}

SCIRL a besoin d'une estimée $\hat{\mu}^{\pi_E}$ de l'attribut moyen de l'expert. C'est un problème similaire à une évaluation de politique. Répétons l'observation-clef : chaque composante de 
$\mu^{\pi_E}$ est une fonction de qualité pour $\pi_E$ vis-à-vis d'une fonction de récompense $\phi_i$: $\mu_i^{\pi_E}(s,a) = Q^{\pi_E}_{\phi_i}(s,a) =
[T^a_{\phi_i} v^{\pi_E}_{\phi_i}](s)$. Nous présentons une revue rapide de méthodes de calcul exacte et approchées ainsi qu'une heuristique.

Si le modèle est connu, il est possible de calculer l'attribut moyen de manière exacte. Soit $\Phi\in\mathbb{R}^{|\s|\times p}$ la matrice d'attributs donc les lignes contiennent les vecteurs d'attributs $\phi(s)^\top$ pour tout
$s\in\s$.
%
%(equivalently, whose columns contain the basis functions $\phi_i$
%for $1\leq i\leq p$).
%
Pour un certain $a\in A$, soit $\Mu^{\pi_E}_a \in\mathbb{R}^{|\s|\times
p}$ la matrice des attributs moyens dont les lignes sont les attributs moyens de l'expert, c'est-à-dire  $(\mu^{\pi_E}(s,a))^\top$ pour chaque $s\in\s$.
Ces notations nous permettent d'écrire $\Mu_a^{\pi_E} = \Phi + \gamma
P_a(I-\gamma P_{\pi_E})^{-1} \Phi$.
%\begin{equation}
%  \forall a\in\A,\; \Mu_a^{\pi_E} = \Phi + \gamma P_a(I-\gamma
%  P_{\pi_E})^{-1} \Phi.
%\end{equation}
%Therefore, knowing the model (the dynamic), the expert feature
%expectation can be easily computed.
Ajoutons que le coût computationel de cette méthode est du même ordre de grandeur que l'évaluation d'une seule politique, en effet la partie coûteuse (le calcul de $(I-\gamma P_{\pi_E})^{-1}$) est partagée par toutes les composantes.

Si le modèle est inconnu, tous les algorithmes d'apprentissage par différences temporelles peuvent être utilisés pour obtenir une estimation de l'attribut moyen de l'expert~\cite{Klein:2011}, comme par exemple LSTD (Least-Squares Temporal
Differences)~\cite{Bradtke:1996}. Soit $\psi:\s\times \A \rightarrow
\mathbb{R}^d$ un vecteur d'attributs composé de $d$ fonctions de base
$\psi_i \in\mathbb{R}^{\s\times \A}$. Chaque composante $\mu_i^{\pi_E}$
 de l'attribut moyen de l'expert est paramétrisée par un vecteur
$\xi_i\in\mathbb{R}^d$: $\mu_i^{\pi_E}(s,a)\approx \xi_i^\top
\psi(s,a)$. Supposons que l'on dispose d'un set d'entraînement
$\{(s_i,a_i,s'_i,a'_i=\pi_E(s'_i))_{1\leq i \leq M}\}$ dont les actions 
$a_i$ ne sont pas nécessairement échantillonnées selon la politique $\pi_E$
(\textit{e.g.}, on peut par exemple utiliser des trajectoires obtenues par un agent suivant une politique $\epsilon$-greedy), le but est d'obtenir une meilleure représentativité des données (des actions sous optimales devraient être essayées). Soit $\tilde{\Psi}\in \mathbb{R}^{M\times d}$
(resp. $\tilde{\Psi}'$) la matrice d'attributs dont les lignes sont les vecteurs d'attributs $\psi(s_i,a_i)^\top$ (resp. $\psi(s'_i,a'_i)^\top$).
Soit $\tilde{\Phi}\in \mathbb{R}^{M\times p}$ la matrive d'attributs dont les lignes sont les vecteurs d'attributs de la récompense $\phi(s_i)^\top$.
Enfin, soit $\Xi =
\begin{bmatrix}   \xi_1 & \dots & \xi_p
\end{bmatrix}\in\mathbb{R}^{d\times p}$ la matrice de tous les vecteurs de paramètres. Appliquer LSTD sur chacune des composantes de l'attribut moyen donne l'algorithme LSTD-$\mu$~\cite{Klein:2011}: $\Xi =
(\tilde{\Psi}^\top(\tilde{\Psi} - \gamma
\tilde{\Psi}'))^{-1}\tilde{\Psi}^\top \tilde{\Phi}$ et
$\hat{\mu}^{\pi_E}(s,a) = \Xi^\top \psi(s,a)$.
%\begin{equation}
%  \Xi = \left(\tilde{\Psi}^\top(\tilde{\Psi} - \gamma
%  \tilde{\Psi}')\right)^{-1}\tilde{\Psi}^\top \tilde{\Phi}
%  \text{ et } \hat{\mu}^{\pi_E}(s,a) = \Xi^\top \psi(s,a).
%\end{equation}
De la même manière que pour le cas exact, la partie coûteuse de l'algorithme (inverser la matrice) est partagée par toutes les composantes. Le coût reste donc raisonnable (du même ordre que LSTD).

%
Si l'on dispose d'un simulateur permettant d'échantillonner selon la politique de l'expert, l'attribut moyen de l'expert peut également être estimé via une méthode de Monte-Carlo pour chaque couple état-action (comme signalé en Sec.~\ref{subsec:scirl:analysis}, $\hat{\mu}^{\pi_E}$ ne doit être connu que pour $(s_i,a)_{1\leq i\leq N,a\in\A}$). Si $K$
trajectoires sont échantillonnées pour chaque couple, cette méthode requiert $KN|\A|$ simulations.

Pour minimiser l'erreur $\bar{\epsilon}_Q$, il est possible d'utiliser des transitions dont l'état de départ est tiré selon 
$\rho_E$ et dont les actions sont uniformément distribuées. Cependant parfois seules les transitions issues de l'expert sont disponibles : $\T =
\{(s_i,a_i=\pi_E(s_i),s'_i)_{1\leq i \leq N}\}$. Bien que le couple état-action $(s_i,a_i)$ puisse être exploité par le classifieur, les transitions $(s_i,a_i,s'_i)$ ne sont pas seules suffisantes pour une estimation précise de l'attribut moyen. Il est toujours possible de rester précis sur l'estimation de $\mu^{\pi_E}(s,\pi_E(s))$, mais il y a peu d'espoir de l'être pour $\mu^{\pi_E}(s,a\neq\pi_E(s))$. Il est heureusement possible d'utiliser une heuristique; ce cas ne rentre pas dans l'analyse présentée en 
Sec.~\ref{subsec:scirl:analysis}, mais peut malgré tout fournir de bons résultats expérimentaux comme illustré en Sec.~\ref{sec:experiments}.

Nous proposons une telle heuristique. Supposons que  $\T$ soient les seules données disponibles, que nous utilisons pour fournir une estimation (précise)
$\hat{\mu}^{\pi_E}(s,\pi_E(s))$ (cela revient à estimer non plus une fonction de qualité comme décrit ci-dessus, mais simplement une fonction de valeur). Un point de vue optimiste suppose que choisir une action différente de celle de l'expert ne fait que retarder l'effet de l'action de l'expert. Plus formellement, nous associons à chaque état $s$ un état virtuel $s_\text{v}$ pour lequel $p(.|s_\text{v},a)=p(.|s,\pi_E(s))$ pour toute action $a$ et pour lequel l'attribut (de récompense) moyen est le vecteur nul, $\phi(s_v) = 0$. Dans ce cas, on a
$\mu^{\pi_E}(s,a\neq\pi_E(s)) = \gamma \mu^{\pi_E}(s,\pi_E(s))$.
Appliquer cette idée sur l'estimation effectivement disponible (rappelons que le classifieur n'a besoin d'évaluer  $\hat{\mu}^{\pi_E}$ qu'en
$(s_i,a)_{1\leq i\leq N,a\in \A}$) fournit l'heuristique que nous proposons :
pour $1\leq i\leq N$, $\hat{\mu}^{\pi_E}(s_i,a\neq a_i) = \gamma
\hat{\mu}^{\pi_E}(s_i,a_i)$.
%\begin{equation}
%  \forall 1\leq i \leq N,\;
%  \hat{\mu}^{\pi_E}(s_i,a\neq a_i) = \gamma
%  \hat{\mu}^{\pi_E}(s_i,a_i).
%  \label{eq:heuristic}
%\end{equation}

Il est également possible de pousser cette idée plus loin afin d'obtenir une estimation plus simple (mais offrant de moindres garanties) de l'attribut moyen de l'expert.
Supposons que  $\T$ consiste en une longue trajectoire, c'est-à-dire
$s'_i = s_{i+1}$ (donc $\T =
\{s_1,a_1,s_2,\dots,s_{N-1},a_{N-1},s_N,a_N\}$). On peut estimer
$\mu^{\pi_E}(s_i,a_i)$ en utilisant la seule trajectoire disponible et en utilisant l'heuristique précedente pour les autres actions :
\begin{equation}
  \forall 1\leq i \leq N,\; \hat{\mu}^{\pi_E}(s_i,a_i) =
  \sum_{j=i}^N \gamma^{j-i}\phi(s_j) \text{ et }
  \hat{\mu}^{\pi_E}(s_i,a\neq a_i) = \gamma
  \hat{\mu}^{\pi_E}(s_i,a_i).
  \label{eq:mc_plus_heuristic}
\end{equation}

Pour résumer, l'attribut moyen de l'expert peut être vu comme un vecteur de fonctions de qualité (pour une même politique  $\pi_E$ et pour différentes fonctions de récompense $\phi_i$). En conséquence, tout algorithme d'évaluation de la fonction de qualité peut être utilisé pour estimer $\mu^\pi(s,a)$.
Selon la quantité et la nature des données disponibles, l'on peut avoir recours à une heuristique pour évaluer l'attribut moyen pour une action différente de celle de l'expert. Cette estimation n'est nécessaire que pour entraîner les classifieur, il est donc suffisant de disposer de valeurs uniquement pour les couples état-action $(s_i,a)_{1\leq i \leq N,a\in \A}$.
En tous cas, estimer  $\mu^{\pi_E}$ n'est pas plus coûteux que d'estimer la fonction de qualité d'une politique donnée, dans le cas \emph{on-policy}, ce qui est rappelons-le moins coûteux que de trouver la politique optimale pour une fonction de récompense arbitraire (comme l'exigent la plupart des algorithmes d'ARI existants, voir Sec.~\ref{sec:relatedWorks}).

\subsection{Instanciation}
\label{subsec:practicalApproach:instantiation}

Comme précisé précedemment, tout algorithme de MC$^2$ peut être utilisé. Ici nous choisissons l'approche à marges structurées de ~\cite{Taskar:2005}. Soit
$\lc:\s\times\A\rightarrow\mathbb{R}_+$ une fonction de marge définie par l'utilisateur satisfaisant $\lc(s,\pi_E(s))\leq \lc(s,a)$ (ici,
$\lc(s_i,a_i)=0$ et $\lc(s_i,a\neq a_i)=1$). L'algorithme MC$^2$ résoud :% the following quadratic problem:
\begin{equation}
  \min_{\theta,\zeta}\frac{1}{2}\|\theta\|^2 +
  \frac{\eta}{N}\sum_{i=1}^N \zeta_i \text{~~~~s.t.~~~~} \forall i,
  \theta^\top\hat{\mu}^{\pi_E}(s_i,a_i)+\zeta_i \geq \max_a \theta^\top
  \hat{\mu}^{\pi_E}(s_i,a) + \lc(s_i,a). \label{eq:qp_taskar}
\end{equation}
Comme dans~\cite{Ratliff:2006}, nous fournissons la forme \emph{hinge-loss} équivalente (avec les variables d'ajustement $\zeta_i$ serrées, ce qui permet de déplacer les contraintes dans la fonction objectif) :
\begin{equation}
  J(\theta) = \frac{1}{N}\sum_{i=1}^N \max_a \theta^\top
  \hat{\mu}^{\pi_E}(s_i,a) + \lc(s_i,a) -
  \theta^\top\hat{\mu}^{\pi_E}(s_i,a_i) +
  \frac{\lambda}{2}\|\theta\|^2.
\end{equation}
La fonction objectif est minimisée en utilisant une descente de sous-gradient. L'attribut moyen de l'expert est estimé en utilisant le principe décrit Eq.~\eqref{eq:mc_plus_heuristic}.
